---
title: "Final Project - Step 3"
author: "Kristal Lee"
date: "2024-05-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Term Project

## 1. Introduction

In the city of Seattle, crime and policing can be a hot topic at times. Media 
can conflate or misrepresent crime rates in the area. As someone who is married 
to a law enforcement officer in Seattle and as someone who is a transplant to the area, 
it would be helpful to know more about the crime in Seattle. What gets reported 
on different media outlets some times paint contradicting pictures of crime and policing in 
Seattle. Personally, I would also like to see a more objective analysis, as I 
may have my own bias along with what I am being told second-hand about crimes in Seattle from my 
husband.

The main purpose of this research paper is to provide a more objective analysis on crime rates in Seattle based off of
data publicly provided by Seattle Police Department. 

## 2. Research Questions
Research questions to explore:  

- Has the amount of crimes changed over the last few year?
  - If so, how by how much?  
  - Within those crimes, has the trend in types of crimes changed?  
- Are there trends or change in trends with the various crimes among different 
areas within Seattle (Precincts/Sectors)?  
- How does use of force and officer involve shootings (OIS) play into this?  
  - Are there more or less over the past few years?  
- What variables that are reported for arrests have greater impact?  
  - Similarly, for any use of force incidents.   

## 3. Approach

Firstly, I would look to see what years overlap within the data sets. I would 
want to shorten the amount of rows because some of the data sets go back over a 
decade with over a million rows. It would make for quicker loading and 
manipulations. Once I fully identify what columns I plan on using, I would also 
remove columns not being utilized. I would use the incident number to connect 
across different data sets for additional information such as if there was an 
OIS or use of force. This would be the main column to join sets. I would also use demographic informations 
to find if there are any correlations with arrests or those who commit crimes.

I would use the combined sets to conduct different types of analysis about the proposed 
research questions. This will be in the form of statistics, tables, and visualizations. 

### 3.1 How Approach Addresses the Problem

The approach will help address the problem by using the direct data from Seattle Police Department (SPD) 
to counter or further support what media has claimed regarding crime in Seattle. 

## 4. Data

All of the data sets that could be used for the analysis are publicly provided on the city of Seattle's government website 
where the data is from SPD. All sets are provided as links to the original site for download. 
Some of the columns that allow for combining the data include incident numbers, location (precincts and sectors), and report numbers for some. 

- [SPD Crime Data: 2008-Present](https://data.seattle.gov/Public-Safety/SPD-Crime-Data-2008-Present/tazs-3rd5/about_data). 
- [SPD Arrest Data](https://data.seattle.gov/Public-Safety/SPD-Arrest-Data/9bjs-7a7w/about_data). 
- [SPD Officer Involved Shooting (OIS) Data](https://data.seattle.gov/Public-Safety/SPD-Officer-Involved-Shooting-OIS-Data/mg5r-efcm/about_data). 
- [Use of Force](https://data.seattle.gov/Public-Safety/Use-Of-Force/ppi5-g2bj/about_data)

```{r libraries}
library(dplyr)
library(purrr)
library(magrittr)
library(ggplot2)
```

```{r load data}
crimedata_loc <- "/Users/kristallee/Documents/Spring_2024/520-Project/SPD_Crime_Data__2008-Present_20240510.csv"
UOF_loc <- "/Users/kristallee/Documents/Spring_2024/520-Project/Use_Of_Force_20240510.csv"
OIS_loc <- "/Users/kristallee/Documents/Spring_2024/520-Project/SPD_Officer_Involved_Shooting__OIS__Data_20240510.csv"

crimedata <- read.csv(crimedata_loc)
UOFdata <- read.csv(UOF_loc)
OISdata <- read.csv(OIS_loc)

```

```{r data samples}
head(crimedata, n=5)
head(UOFdata, n=5)
head(OISdata, n=5)
```

```{r years in df}
# add year column from date column for all datasets
crimedata$Year <- format(as.Date(crimedata$Offense.Start.DateTime, 
                                 format="%m/%d/%Y"), "%Y")
UOFdata$Year <- format(as.Date(UOFdata$Occured_date_time, 
                               format="%m/%d/%Y"), "%Y")
OISdata$Year <- format(as.Date(OISdata$Date...Time, format="%m/%d/%Y"), "%Y")


```
```{r crime counts by year, all}
# bar graphs for each dataset for all years prior to current with suffient
# amount of data
allcrimecount_year<-crimedata %>% dplyr::count(Year)
allcrimecount_year <- rename(allcrimecount_year, count = n)
allcrimecount_year <- filter(allcrimecount_year, Year>=2009 & Year<2024)
ggplot(allcrimecount_year, aes(x=Year, y=count)) + 
                          geom_bar(stat="identity") + ggtitle("Crime Reported by Year")
allUOFcount_year<-UOFdata %>% dplyr::count(Year)
allUOFcount_year <- rename(allUOFcount_year, count = n)
allUOFcount_year <- filter(allUOFcount_year, Year<2024)
ggplot(allUOFcount_year, aes(x=Year, y=count)) + 
                          geom_bar(stat="identity") + ggtitle("UOF by Year")
allOIScount_year<-OISdata %>% dplyr::count(Year)
allOIScount_year <- rename(allOIScount_year, count = n)
allOIScount_year <- filter(allOIScount_year, Year<2024)
ggplot(allOIScount_year, aes(x=Year, y=count)) + 
                          geom_bar(stat="identity") + ggtitle("OIS by Year")
                        
```

```{r unique years}
# find most recent years that overlap
crimeyears <- unique(sort(crimedata$Year, decreasing=TRUE))
UOFyears <- unique(sort(UOFdata$Year, decreasing=TRUE))
OISyears <- unique(sort(OISdata$Year, decreasing=TRUE))

crimeyears
UOFyears
OISyears

# Most recent overlapping 4 year period, 2019-2022
```

```{r overlapping years datasets}
# create shortened datasets for data years that overlap (2018-2022)
newcrimedata <- crimedata %>% filter(Year>= 2019 & Year <= 2022)
newUOFdata <- UOFdata %>% filter(Year>= 2019 & Year <= 2022)
newOISdata <- OISdata %>% filter(Year>= 2019 & Year <= 2022)

# check the unique years 
unique(sort(newcrimedata$Year))
unique(sort(newUOFdata$Year))
unique(sort(newOISdata$Year))
```

```{r months}
# add months column from date column for possibly monthly breakdown
newcrimedata$Month <- format(as.Date(newcrimedata$Offense.Start.DateTime, 
                                 format="%m/%d/%Y"), "%m")
newOISdata$Month <- format(as.Date(newOISdata$Date...Time, 
                                   format="%m/%d/%Y"), "%m")
newUOFdata$Month <- format(as.Date(newUOFdata$Occured_date_time, 
                               format="%m/%d/%Y"), "%m")
```


```{r counts by year}
# create tables and plots for number of crimes 
countscrimes <-table(newcrimedata$Year)
barplot(countscrimes, main="Crimes by Year", xlab="years")
countsUOF <-table(newUOFdata$Year)
barplot(countsUOF, main="UOF by Year", xlab="years")
countsOIS <-table(newOISdata$Year)
barplot(countsOIS, main="OIS by Year", xlab="years")

```

```{r top ten groups}
# unique(newcrimedata$Offense.Parent.Group)
offenseCount<-newcrimedata %>% dplyr::count(Offense.Parent.Group)
offenseCount <- rename(offenseCount, count = n)
top_10 <- slice_max(offenseCount, order_by = count, n=10)
```

```{r top ten df}
top_10_crime<- newcrimedata %>% 
  filter(Offense.Parent.Group==top_10$Offense.Parent.Group)
test <- table(top_10_crime$Offense.Parent.Group, top_10_crime$Year)
top10.df <- data.frame(test)
colnames(top10.df) <- c('Offense.Parent.Group', 'Year', 'Count')
top10.df
```

```{r plot top 10}
top10_plot <- ggplot(top10.df, aes(fill=Offense.Parent.Group, y=Count, x=Year)) + geom_bar(position="fill", stat="identity")
print(top10_plot + labs(title="Ratio of Top 10 Offense Groups by Year", y="Percentage", x="Year"))
```
```{r plot top precints}
precinctCount <-newcrimedata %>% dplyr::count(Precinct)
precinctCount <- rename(precinctCount, count = n)
precinctCount

```


```{r plot top precints2}
test2 <- newcrimedata %>% filter(Precinct != "OOJ") %>% group_by(Precinct) %>% 
                                                 filter(Offense.Parent.Group == top_10$Offense.Parent.Group)
test3 <- table(test2$Offense.Parent.Group, test2$Precinct)
precTop10.df <- data.frame(test3)
colnames(precTop10.df) <- c('Offense.Parent.Group', 'Precinct', 'Count')
```

```{r plot top precints3}
precplot <- ggplot(precTop10.df, aes(x=Precinct, y=Count, fill=Precinct)) + geom_bar(stat="identity") + facet_wrap(~Offense.Parent.Group) + theme(axis.title.x=element_blank(), axis.text.x=element_blank(),axis.ticks.x=element_blank())
print(precplot + labs(title="Top Ten Reported Crimes by Precinct (2019-2022)", y="Count"))

```
```{r clean UOF}
exclude_p <- list("0", "-" ,"OOJ")
newUOFdata <- newUOFdata %>% filter(! Precinct %in% exclude_p)
newUOFdata <- newUOFdata %>%
  mutate(Precinct = recode(Precinct, East = 'E', North = 'N', South = 'S', West = 'W', Southwest = 'SW'))
unique(newUOFdata$Precinct)
```
```{r UOF plot}
countsUOF_bypre <-table(newUOFdata$Precinct)
barplot(countsUOF_bypre, main="UOF by Precinct", xlab="Precincts")

```


